---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# APPENDIX R-CODE

<!-- badges: start -->
<!-- badges: end -->

The goal of the Appendix of R-code is to provide interested readers in more information about the Bayesian model and modeling code.

This document contains more information on the Bayesian estimation and statistical analysis. The main analysis script (`main_script.R`) contains the R-code that was used to load and restructure the data. The Bayesian model as written in stan can be consulted in the `stan_model.stan`-file.

## Bayesian methods

In this paper, our goal is to estimate the true underlying seroprevalence of the population $\geq 18$  and $\leq 65$ years old as measured every two weeks, $w$ in Belgium, denoted $p^*_w$ ($w=1,\dots,W=20$).

We start by estimating the probability that each person in the serosurvey is seropositive using a Bayesian logistic regression model that accounts for the sensitivity and specificity of the ELISA assay, each individual's age, sex, and current province, iterated over each two weeks they were sampled:

\begin{align*}
  x_{i} &\sim Bernoulli(p_i \theta^+ + (1-p_i) * (1-\theta^-)) \\
  logit(p_i) &= \alpha_h + \mathbf{X}_i \boldsymbol{\beta} \\
  \alpha_h &\sim Normal(0, \sigma^2) \\
  x^+ &\sim Binomial(n^+, \theta^+) \\
  x^- &\sim Binomial(n^-, 1-\theta^-)
\end{align*}

where $x_i$ is the result of the IgG ELISA (in primary analyses) for the $i$th person ($i=1,\dots,N=$ `r nrow(df)`) in the serosurvey.
The sensitivity, $\theta^+$, is determined using $n^+$ RT-PCR positive controls from the lab validation study, of which $x^+$ tested positive.
The specificity, $\theta^-$ , is determined using $n^-$ pre-pandemic negative controls, of which $x^-$ tested positive.

The probability of observing a diagnostic positive is a function of the true positive rate and the false negative rate with regards to the true underlying probability of seropositivity $p_i$ for that person.
This probability itself is a function of covariates $\mathbf{X}$, which consists of sex, age categories, and week of study, and their coefficients $\boldsymbol{\beta}$, and a random effect for household, $\alpha_h$ ($h=1,\dots, H=$ `r nrow(df)`), with variance $\sigma^2$.
We used naive priors on all parameters to allow for an exploration of the parameter space.
The priors on the sensitivity and specificity were flat from 0 to 1, equivalent to $Uniform(0,1)$ or $Beta(1,1)$.
We used weak $Normal(0,3)$ priors for the logistic regression coefficients $\boldsymbol{\beta}$.
The prior on the standard deviation of the household effect, $\sigma$, was flat from 0 to infinity (we tested a positive half-Normal and it did not affect estimates).

## Bayesian MCMC Diagnostics

### Divergence

In a first step, the number of chains that diverged where evaluated using the `rstan::check_divergences`-function:
`r rstan::check_divergences(fit)`

### Trace plots

A first diagnostic plot is the trace plot, which is a time series plot of the Markov chains. That is, a trace plot shows the evolution of parameter vector over the iterations of one or many Markov chains

```{r quality}
## check the model
posterior_fit <- as.array(fit) ## save as array to plot
pl_age <- mcmc_trace(posterior_fit, regex_pars = c("age"))
pl_sex <- mcmc_trace(posterior_fit, regex_pars = c("sex"))
pl_prov <- mcmc_trace(posterior_fit, regex_pars = c("prov"))
```

For age, this resulted in the following trace plot:

```{r trace_age, fig.width=6}
pl_age
```

For sex, this resulted in the following trace plot:

```{r trace_sex, fig.width=6}
pl_sex
```

For province, this resulted in the following trace plot:

```{r trace_prov, fig.width=6}
pl_prov
```

### R-hat

Another way to evaluate chain convergence to an equilibrium distribution is to compare its behavior to other  initialized chains. This is the motivation for the potential scale reduction statistic, split-$\hat{R}$. The split-$\hat{R}$ statistic measures the ratio of the average variance of draws within each chain to the variance of the pooled draws across chains; if all chains are at equilibrium, these will be the same and split-$\hat{R}$ will be one. If the chains have not converged to a common distribution, the split-$\hat{R}$ statistic will be greater than one (see Gelman et al. 2013, Stan Development Team 2018).

```{r rhat}
## RHAT STATISTIC
rhats <- rhat(fit, pars = c("beta_age", "beta_sex", "beta_prov", 
                                            "sigma_age", "sigma_prov", "sigma_sex"))
mcmc_rhat(rhats)
```

### ESS

The effective sample size is an estimate of the number of independent draws from the posterior distribution of the estimand of interest. The $n_{eff}$ metric used in Stan is based on the ability of the draws to estimate the true mean value of the parameter. Because the draws within a Markov chain are not independent if there is autocorrelation, the effective sample size, $n_{eff}$, is usually smaller than the total sample size, N. The larger the ratio of $n_{eff}$ to N the better (see Gelman et al. 2013, Stan Development Team 2018 for more details) .

```{r ess}
## ESS STATISTIC
ess <- neff_ratio(fit, pars = c("beta_age", "beta_sex", "beta_prov", 
                                                "sigma_age", "sigma_prov", "sigma_sex",
                                                "phi"))
mcmc_neff(ess, size = 2)  
```

## Bayesian Output

Bayesian parameter estimation was performed by calculating the median of the posterior draws of $\phi$. The 95% CI was calculated based on the Highest (Posterior) Density Interval (HDI) of $\phi$.

```{r parest}
est <- tibble(median = median(fit_mcmc$phi),
              hdill = hdi(fit_mcmc$phi, credMass = 0.95)[1],
              hdiul = hdi(fit_mcmc$phi, credMass = 0.95)[2])
```


